---
title: 'Final Project: Navajo Nation Water Quantity and Quality Analysis from 2019-2024'
author: "Samantha White-Murillo, Wynona Curaming, Siyu Dong" 
date: "Spring 2024"
output: 
  pdf_document:
   fig_caption: true
---
\newpage
\tableofcontents
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Working directory
getwd()

# Load your packages
library(dataRetrieval)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)
library(here)
library(ggthemes)
library(pdftables)
library(viridis)
library(rvest)
library(tidycensus)
library(ggplot2)

# Set your ggplot theme
my_theme <- theme_light() +
  theme(plot.title = element_text(face = "bold", size = 12, color = "black"),
        axis.title = element_text(color = "darkblue", )) +
  theme(legend.position = "bottom")

# Load your data sets
## Water Quality
Boron_Site1 <- read.csv("./Raw_data/Boron_S1.csv", stringsAsFactors = TRUE)
Boron_Site2 <- read.csv("./Raw_data/Boron_S2.csv", stringsAsFactors = TRUE)
Magnesium_Site1 <- read.csv("./Raw_data/magnesium_S1.csv", stringsAsFactors = TRUE)
Magnesium_Site2 <- read.csv("./Raw_data/magnesium_S2.csv", stringsAsFactors = TRUE)
Uranium_Site1 <- read.csv("./Raw_data/Uranium_S1.csv", stringsAsFactors = TRUE)
Uranium_Site2 <- read.csv("./Raw_data/Uranium_S2.csv", stringsAsFactors = TRUE)

## Water Quantity 
siteNumber <- "09379500"
ChoptankInfo <- readNWISsite(siteNumber)
parameterCd <- "00060"

# Raw daily data:
sanjuan_quantity_df <- readNWISdv(
  siteNumber, parameterCd,
  "2019-04-24", "2024-04-22"
)

pCode <- readNWISpCode(parameterCd)

siteNumber2<- "09380000"
ChoptankInfo2 <- readNWISsite(siteNumber2)
parameterCd2 <- "00060"

# Raw daily data:
colorado_quantity_df <- readNWISdv(
  siteNumber2, parameterCd2,
  "2019-04-24", "2024-04-22"
)

pCode2 <- readNWISpCode(parameterCd2)

```


# Rationale and Research Questions
The Navajo Nation has severe water infrastructure deficiencies that impact the health, economy, and welfare of the Navajo people. The lack of adequate domestic and municipal water is the greatest water resource problem facing the Navajo Nation. Approximately 30-40% of the Navajo Nation population does not have access to clean, reliable drinking water  (Navajo Nation Department of Water Resources, 2024). Also, 173 thousand people are affected because drinking water sources are limited, and abandoned uranium mines have caused groundwater contamination in the Nation (Ibid.). In addition, many improvements are needed for other areas of water use, including water for irrigation, livestock, commercial, businesses, health care, schools, and other facilities (SOURCE, n.d.).

In terms of water quantity, climate change in the Southwest will continue to impact water resources problems. The USGS Disaster Risk Assessment Study concluded that a long-term drying trend and decreasing snowpack, superimposed on the regional drought cycles, will magnify water-related impacts in the Navajo Nation and leave the Navajo people increasingly vulnerable (Navajo Nation Department of Water Resources, 2024). 

The current project focuses on analyzing the water quality of the main Colorado River before and after it flows through the reservation over the past 5 years. A five-year analysis period was selected to provide a more current evaluation. Taking into account the tribe’s persistent efforts to compel the federal government to meet its obligations, aiding in the quantification of the tribe’s water rights on the Colorado River and ensuring access to high-quality water, we have chosen to focus our quality analysis on the Colorado River. This decision stems from the river’s current water rights litigation and data availability. This analysis aims to test the following hypotheses: 

> Hypothesis A: The water quality in the Colorado River before and after passing through the Navajo Nation is significantly different.

> Hypothesis B: The water quality in the Colorado River before and after passing through the Navajo Nation has changed significantly in the last 5 years.

On the other hand, measurements have been conducted for the water quantity of the rivers over the last 5 years (2019-2024) to assess water availability in both the Colorado River and the San Juan River. In order to broaden the scope of our analysis of water availability, the San Juan River was also selected, establishing the following hypothesis:

> Hypothesis C: Water availability changed in the last years for the Navajo Nation.


# Dataset Information
The dataset was taken from the USGS Monitoring the rivers of the Nation (https://waterdata.usgs.gov/nwis).

For water quality analysis, data was extracted from the years 2018 to 2023, capturing Uranium, Magnesium, and Boron levels detected at both stations. The reason for selecting these components was the availability of data at both stations. It was necessary for both stations to have the same components during the same period to enable a comparative analysis. Also, those component represent a high risk for human health and the environment at high levels [***I think we should elaborate with one or two sentences and cite].

Below is the information on the stations worked:
- Station 1 = Colorado River at Lees Ferry, AZ - 09380000
- Station 2 = Colorado River Near Grand Canyon, AZ - 09402500

For water quantity, the data was extracted from years 2019 to 2024 from station 1 as well, extracting data of water flow in cf/s.

Since there is no missing data it is not necessary to execute linear interpolation for time series analysis. 

\newpage

# Exploratory Analysis 

## Water Quantity
#### Question 1: How have water discharge levels changed in the last 5 years at San Juan and Colorado rivers?


```{r Data Wrangling, echo=FALSE}
#Renaming columns to be understandable 
sanjuan_quantity_df<-rename(sanjuan_quantity_df, discharge_cf = X_00060_00003)
colorado_quantity_df<-rename(colorado_quantity_df, discharge_cf = X_00060_00003)

```

### Data Visualization

The observed downward trend in discharge volume, as evidenced by measurements from both gages (Figures 1 and 2), presents a notable concern regarding the hydrological health of the San Juan and Colorado Rivers over the preceding half-decade. This declining trend signifies a significant challenge to the sustainability of these crucial aquatic ecosystems.

Of particular interest are changes in discharge levels in the San Juan River. While both rivers exhibit an overall decrease in discharge, the San Juan demonstrates more frequent instances of elevated discharge levels. Noteworthy spikes in discharge, notably observed around the midpoints of 2023 and 2019, highlight the river's seasonal variability and susceptibility to external influences.

```{r Data Visualization San Juan, fig.align='center', fig.cap="Water Levels in San Juan River", message=FALSE, echo=FALSE}
ggplot(data=sanjuan_quantity_df, 
       aes(x=Date, y=discharge_cf)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(y= "Discharge (cubic feet/second)") +
  my_theme

```

```{r Data Visualization Colorado, echo=FALSE, fig.align='center', fig.cap="Water Levels in Colorado River", message=FALSE}
ggplot(data = colorado_quantity_df, 
       aes(x=Date, y=discharge_cf)) +
  geom_point()+geom_smooth(method="lm") +
  labs(y= "Discharge (cubic feet/second)") +
  my_theme

```

### Time Series Analysis

For Colorado River, water levels have declined from 2019 to mid-2022 and then increased to peak at early 2023 (Figure 4). For San Juan River, water levels have declined from 2019 to 2020 and remained steady (at a low level) up till mid-2022. Suddenly, water levels increased to peak at early 2023, and declined since then (Figure 5).

```{r Time Series Analysis San Juan River, echo=FALSE, fig.align='center', fig.cap="Time Series Analysis of San Juan River, 2019-2024"}

# San Juan River 
# Generate time series

f_month <- month(first(sanjuan_quantity_df$Date))
f_year <- year(first(sanjuan_quantity_df$Date))
f_day<- day(first(sanjuan_quantity_df$Date))

l_month <- month(last(sanjuan_quantity_df$Date))
l_year <- year(last(sanjuan_quantity_df$Date))
l_day<- day(last(sanjuan_quantity_df$Date))

sanjuan_quantity_daily_ts <- ts(sanjuan_quantity_df$
                              discharge_cf,
                            start=c(f_year,f_month),
                            frequency=365)

# Decompose

sanjuan_quantity_daily_decomposed <- stl(sanjuan_quantity_daily_ts, s.window = "periodic")
plot(sanjuan_quantity_daily_decomposed)

```

```{r Time Series Analysis Colorado River, echo=FALSE, fig.align='center', fig.cap="Time Series Analysis of Colorado River, 2019-2024"}

# Colorado River  
# Generate time series
f_month1 <- month(first(colorado_quantity_df$Date))
f_year1 <- year(first(colorado_quantity_df$Date))
f_day1<- day(first(colorado_quantity_df$Date))

l_month1 <- month(last(colorado_quantity_df$Date))
l_year1 <- year(last(colorado_quantity_df$Date))
l_day1<- day(last(colorado_quantity_df$Date))

colorado_quantity_daily_ts<- ts(colorado_quantity_df$
                              discharge_cf,
                            start=c(f_year,f_month),
                            frequency=365)

# Decompose
colorado_quantity_daily_decomposed<- stl(colorado_quantity_daily_ts, s.window = "periodic")
plot(colorado_quantity_daily_decomposed)
```


### Population Drawing Directly From the Rivers 

We examined each of the chapters under the Western and Northern Agency on whether the San Juan and Colorado River passes through their territory (Navajo Nation Division of Community Development, n.d.). This would allows us to determine how much of the Navajo population drew water directly from each river. 

For Colorado River, the following chapters met this criteria: Bodaway-Gap, Coppermine, LeChee, and Navajo Mountain. For San Juan, the following chapters met this criteria: Kayenta, Oljato, Shonto, Beclabito, Upper Fruitland, Gadii Ahi/to'koi, Mexican Water, Red Mesa, Teec Nos Pos, Tse Daa Kaan. 

The total Indian population in the Navajo Nation estimated to live in the chapters bordering the Colorado River is about 4516 (Navajo Nation Division of Community Development, n.d.). San Juan River has about 18728 people living close-by (Ibid.). 

```{r Manual data collection, echo=FALSE}

#Population Directly Drawing Water from the Colorado River  
pop_Bodaway_Gap<-1705
pop_Coppermine<-690
pop_LeChee<-1762
pop_Navajo_Mt<-359
pop_chapters_Colorado<- pop_Bodaway_Gap+pop_Coppermine+pop_LeChee+pop_Navajo_Mt

#Population Directly Drawing Water from the San Juan River 
pop_Kayenta<-6035
pop_Oljato<-2729
pop_Shonto<-1645
pop_Beclabito<-724
pop_Upper_Fruitland<-2594
pop_Gadii_Ahi<-430
pop_Mexican_Water<-667
pop_Red_Mesa<-1177
pop_Teec_Nos_Pos<-1454
pop_Tse_Daa_Kaan <-1273

pop_chapters_SJ<-pop_Kayenta+pop_Oljato+pop_Shonto+pop_Beclabito+pop_Upper_Fruitland+pop_Gadii_Ahi+pop_Mexican_Water+pop_Red_Mesa+pop_Teec_Nos_Pos+pop_Tse_Daa_Kaan

```

*Note: We intended to use scrape data from the website but it generated empty values. Hence, copy pasting the data was done instead. See appendix 1.*

Navajo people are estimated to use only 7 gallons of water per day (Supreme Court of the United States, 2022). Based on this estimate, and the estimate of population drawing water directly from the rivers, we can estimate the gallons of annual water withdrawal by the Navajo Nation. About 11538380 gallons in total are directly withdrawn from the Colorado River by the Navajo Nation, while about 47850040 gallons in total are directly withdrawn by the Navajo Nation from the San Juan River. This is a small quantity compared to [*i think we need a statistic of how much other states withdraw from either river to put these numbers in context]

This estimate applies only for domestic use but since they use water for other activities such as agriculture, the water availability of both rivers is a limiting factor. 

Note: we're counting only populations right beside the river, even though with water transport infrastructure, people in chapters far from the river will also be withdrawing, so this estimate of water withdrawing is an underestimate.  


```{r}
seconds_per_year<-365*24*60*60
#water available in periods of minimum water discharge from the last 5 years
water_available_SJ_2024<-500*7.48052*seconds_per_year
water_available_SJ_2024 
water_available_Colorado_2024<-9500*7.48052*seconds_per_year
water_available_Colorado_2024  

#Note to samantha: Do we need this? Cos even at the lowest 
```

```{r echo=FALSE, message=FALSE}
#Colorado 
annual_draw_colorado<-4516*7*365
annual_draw_colorado
#San Juan 
annual_draw_SJ<-18728*7*365
annual_draw_SJ

```

cubic feet per second
1 cubic feet = 7.48052 gallons 
```{r echo=FALSE, message=FALSE}
seconds_per_year<-365*24*60*60
#water available in periods of minimum water discharge from the last 5 years
water_available_SJ_2024<-500*7.48052*seconds_per_year
water_available_SJ_2024 
water_available_Colorado_2024<-9500*7.48052*seconds_per_year
water_available_Colorado_2024  

#Note to samantha: Do we need to include this in our analysis? Cos even at the lowest levels, there is enough water passing through to meet the minimal 7gallon/day consumption 
```
*make description*

\newpage


## Water Quality
#### Question 2: How has been the water quality of Colorado River before, and after it pass through the reservation in the past 5 years?

```{r Dara Preparation, echo=FALSE, message=FALSE, warning=FALSE}

#Chage Date Format
class(Boron_Site1$Activity_StartDate)
class(Boron_Site2$Activity_StartDate)
class(Magnesium_Site1$Activity_StartDate)
class(Magnesium_Site2$Activity_StartDate)
class(Uranium_Site1$Activity_StartDate)
class(Uranium_Site2$Activity_StartDate)

Boron_Site1$Activity_StartDate <- as.Date(Boron_Site1$Activity_StartDate, format = "%Y-%M-%d")
Boron_Site2$Activity_StartDate <- as.Date(Boron_Site2$Activity_StartDate, format = "%Y-%M-%d")
Magnesium_Site1$Activity_StartDate <- as.Date(Magnesium_Site1$Activity_StartDate, format = "%Y-%M-%d")
Magnesium_Site2$Activity_StartDate <- as.Date(Magnesium_Site2$Activity_StartDate, format = "%Y-%M-%d")
Uranium_Site1$Activity_StartDate <- as.Date(Uranium_Site1$Activity_StartDate, format = "%Y-%M-%d")
Uranium_Site2$Activity_StartDate <- as.Date(Uranium_Site2$Activity_StartDate, format = "%Y-%M-%d")

class(Boron_Site1$Activity_StartDate)
class(Boron_Site2$Activity_StartDate)
class(Magnesium_Site1$Activity_StartDate)
class(Magnesium_Site2$Activity_StartDate)
class(Uranium_Site1$Activity_StartDate)
class(Uranium_Site2$Activity_StartDate)

#Merger Datasets respectivelly
Boron_merged <- bind_rows(Boron_Site1, Boron_Site2)
Magnesium_merged <- bind_rows(Magnesium_Site1, Magnesium_Site2)
Uranium_merged <- bind_rows(Uranium_Site1, Uranium_Site2)

#Unify the research time range
Uranium_filtered <- Uranium_merged %>%
  filter(Activity_StartDate >= as.Date("2018-01-01") & 
         Activity_StartDate <= as.Date("2023-12-31"))

#Generate the annual average concentration of three contaminations for both sites
Boron_merged$Year <- lubridate::year(Boron_merged$Activity_StartDate)
Boron_annual_average <- Boron_merged %>%
  group_by(Year, Location_Name) %>%
  summarise(Avg_Concentration = mean(Result_MeasureValue, na.rm = TRUE))

Magnesium_merged$Year <- lubridate::year(Magnesium_merged$Activity_StartDate)
Magnesium_annual_average <- Magnesium_merged %>%
  group_by(Year, Location_Name) %>%
  summarise(Avg_Concentration = mean(Result_MeasureValue, na.rm = TRUE))

Uranium_filtered$Year <- lubridate::year(Uranium_filtered$Activity_StartDate)
Uranium_annual_average <- Uranium_filtered %>%
  group_by(Year, Location_Name) %>%
  summarise(Avg_Concentration = mean(Result_MeasureValue, na.rm = TRUE))

```

### Data Visualization Analysis

There is an increase on the mean Boron concentrations along the years on the Colorado river (Figure 5), never the less, that increment and the concentrations itself haven't exceed a concerning level for human health according to the recommended maximum levels of drinking water of Canada and The World Health Organization standards (World Health Organization, 2009; Health Canada, 2020). 

In all the time periods there is less concentration of Boron when entering the tribe territory than when it exits, indicating that within the activities of the nation there is boron discharge in the Colorado river.

```{r Data visualization Boron, echo=FALSE, fig.align='center', fig.cap="Boron Concentration before and after passing through the Navajo Nation"}

#Compare Boron
ggplot(Boron_annual_average, 
       aes(x = Year, y = Avg_Concentration, color = Location_Name)) +
  geom_point() +
  labs(x = "Observation Year",
       y = "Mean Concentration (ug/L)") +
  my_theme

```

Magnesium concentrations in the river show. 

```{r Data visualization Magnesium, echo=FALSE, fig.align='center', fig.cap="Magnesium Concentration before and after passing through the Navajo Nation"}

#Compare Magnesium
ggplot(Magnesium_annual_average, 
       aes(x = Year, y = Avg_Concentration, color = Location_Name)) +
  geom_point() +
  labs(x = "Observation Year",
       y = "Mean Concentration (ug/L)") +
  my_theme

```

```{r Data visualization Uranium, echo=FALSE, fig.align='center', fig.cap="Uranium Concentration before and after passing through the Navajo Nation"}

#Compare Uranium
ggplot(Uranium_annual_average, 
       aes(x = Year, y = Avg_Concentration, color = Location_Name)) +
  geom_point() +
  labs(x = "Observation Year",
       y = "Mean Concentration (ug/L)") +
  my_theme

```

\newpage

# Summary and Conclusions


\newpage

# References

Health Canada. (2020). Boron in drinking water. In Guidelines for Canadian Drinking Water Quality: Guideline Technical Document [Report]. https://www.canada.ca/content/dam/hc-sc/documents/programs/consultation-boron-drinking-water/document/boron-gtd-consultation-09-12-2019-eng.pdf

Navajo Nation Division of Community Development (n.d.). Navajo Nation Profile. Navajo Nation WIND. Retrieved April 28, 2024, from https://navajoprofile.wind.enavajo.org/

Navajo Nation Water Rights Comission. (2024, 3 22). Navajo Nation Water Rights Comission 
webpage. Retrieved from https://nnwrc.navajo-nsn.gov/

SOURCE. (n.d.). Los hidropaneles SOURCE llevan agua a la Nación Navajo. Retrieved from https://www.source.co/resources/case-studies/los-hidropaneles-source-llevan-agua-a-lanacion-navajo/?lang=es

Supreme Court of the United States (2022, October 1). Arizona et al. V. Navajo Nation et al. Retrieved April 28, 2024, from https://www.supremecourt.gov/opinions/22pdf/21-1484_aplc.pdf

The Colorado River: Heading into 2024 with hope for a more stable system | Arizona Department of Water Resources. (2023, December 21). https://www.azwater.gov/news/articles/2023-12-21-0

UN-Water Decade Programme on Advocacy and Communication and Water Supply and Sanitation Collaborative Council. (2010). The human right to water and sanitation today. In Media Brief. https://www.un.org/waterforlifedecade/pdf/human_right_to_water_and_sanitation_media_brief.pdf

Williams, A. P., Cook, B. I., & Smerdon, J. E. (2022). Rapid intensification of the emerging southwestern North American megadrought in 2020–2021. Nature Climate Change, 12(3), 232–234. https://doi.org/10.1038/s41558-022-01290-z

World Health Organization. (2009). Boron. Chemical Fact Sheets. https://www.who.int/docs/default-source/wash-documents/wash-chemicals/boron-chemical-fact-sheet.pdf?sfvrsn=a936384f_4



\newpage

# Appendix
## Appendix 1
```{r Data scraping attempt}
#set scraping website 
Water_URL<- read_html('https://navajoprofile.wind.enavajo.org/')
Water_URL

#scrape the data 
navajo_nation_chapter<-Water_URL%>%
  html_nodes("td:nth-child(3)")%>%
  html_text()
navajo_nation_chapter

indian_population<-Water_URL%>%
  html_nodes("td:nth-child(4)")%>%
  html_text()
indian_population


table<-Water_URL%>%
  html_nodes("tabContent001")%>%
  html_text()
table

```
