---
title: 'Final Project: Navajo Nation Water Quality Analysis'
author: "Samantha White-Murillo, Wynnoa, Iby" 
date: "Spring 2024"
output: 
geometry: margin=2.54cm
  pdf_document:
   fig_caption: true
   editor_options: 
  chunk_output_type: console
---
\newpage
\tableofcontents
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Working directory
getwd()

# Load your packages
library(dataRetrieval)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)
library(here)
library(ggthemes)
library(pdftables)
library(viridis)
library(rvest)
library(tidycensus)

# Set your ggplot theme
my_theme <- theme_light() +
  theme(plot.title = element_text(face = "bold", size = 12, color = "black"),
        axis.title = element_text(color = "darkblue", )) +
  theme(legend.position = "bottom")

# Load your datasets
## Water Quality
read_csv("Raw_data/Boron_S1.csv")
read_csv("Raw_data/Boron_S2.csv")
read_csv("Raw_data/magnesium_S1.csv")
read_csv("Raw_data/magnesium_S2.csv")
read_csv("Raw_data/Uranium_S1.csv")
read_csv("Raw_data/Boron_S1.csv")

## Water Quantity 
siteNumber <- "09379500"
ChoptankInfo <- readNWISsite(siteNumber)
parameterCd <- "00060"

# Raw daily data:
sanjuan_quantity_df <- readNWISdv(
  siteNumber, parameterCd,
  "2019-04-24", "2024-04-22"
)

pCode <- readNWISpCode(parameterCd)

siteNumber2<- "09380000"
ChoptankInfo2 <- readNWISsite(siteNumber2)
parameterCd2 <- "00060"

# Raw daily data:
colorado_quantity_df <- readNWISdv(
  siteNumber2, parameterCd2,
  "2019-04-24", "2024-04-22"
)

pCode2 <- readNWISpCode(parameterCd2)

```


# Rationale and Research Questions



\newpage

# Dataset Information



\newpage

# Exploratory Analysis 

## Water Quantity
### Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Data Wrangling
```{r}
#Renaming columns to be understandable 
sanjuan_quantity_df<-rename(sanjuan_quantity_df, discharge_cf = X_00060_00003)
colorado_quantity_df<-rename(colorado_quantity_df, discharge_cf = X_00060_00003)

```

## Data Visualization 
```{r, data exploration}
ggplot(data=sanjuan_quantity_df, aes(x=Date, y=discharge_cf))+geom_point()+geom_smooth(method="lm")
ggplot(data=colorado_quantity_df, aes(x=Date, y=discharge_cf))+geom_point()+geom_smooth(method="lm")

discharge_sanjuan<-sanjuan_quantity_df$discharge_cf
plot(discharge_sanjuan)
```
> The downward slope measured at both gages suggests that discharge volume for both San Juan and Colorado Rivers have declined in the last five years. San Juan appears to have more frequent periods of high discharge than usual (eg. mid-2023 and mid-2019) compared to Colorado. 

## Time Series Analysis 
Study question: How have water discharge levels changed in the last 5 years at San Juan and Colorado rivers?

We don't have any missing data so we don't need to do linear interpolation. 

```{r}
# San Juan River 
# Generate time series
f_month <- month(first(sanjuan_quantity_df$Date))
f_year <- year(first(sanjuan_quantity_df$Date))
f_day<- day(first(sanjuan_quantity_df$Date))

l_month <- month(last(sanjuan_quantity_df$Date))
l_year <- year(last(sanjuan_quantity_df$Date))
l_day<- day(last(sanjuan_quantity_df$Date))

sanjuan_quantity_daily_ts<- ts(sanjuan_quantity_df$
                              discharge_cf,
                            start=c(f_year,f_month),
                            frequency=365)

# Decompose

sanjuan_quantity_daily_decomposed<- stl(sanjuan_quantity_daily_ts, s.window = "periodic")
plot(sanjuan_quantity_daily_decomposed)

```

```{r}
# Colorado River  
# Generate time series
f_month1 <- month(first(colorado_quantity_df$Date))
f_year1 <- year(first(colorado_quantity_df$Date))
f_day1<- day(first(colorado_quantity_df$Date))

l_month1 <- month(last(colorado_quantity_df$Date))
l_year1 <- year(last(colorado_quantity_df$Date))
l_day1<- day(last(colorado_quantity_df$Date))

colorado_quantity_daily_ts<- ts(colorado_quantity_df$
                              discharge_cf,
                            start=c(f_year,f_month),
                            frequency=365)

# Decompose
colorado_quantity_daily_decomposed<- stl(colorado_quantity_daily_ts, s.window = "periodic")
plot(colorado_quantity_daily_decomposed)
```

## Description of plots 

Colorado River: Water levels have declined from 2019 to mid-2022, then increased to peak at early 2023. 

San Juan River: Water levels have declined from 2019 to 2020 and remained steady (at a low level) up till mid-2022. Water levels then increased to peak at early 2023, and declined since then. 

## Population Drawing Directly From the Rivers 

check how much water the Navajo people withdraw or need?
check population living near the river-> est how much they would need 
potential explanation- agriculture

We examined each of the chapters under the Western and Northern Agency on whether the San Juan and Colorado River passes through their territory. This would allows us to determine how much of the Navajo population collected water from each respective river. 

For the Colorado River, the following chapters met this criteria: Bodaway-Gap, Coppermine, LeChee, and Navajo Mountain. 

For San Juan, the following chapters met this criteria: Kayenta, Oljato, Shonto, Beclabito, Upper Fruitland, Gadii Ahi/to'koi, Mexican Water, Red Mesa, Teec Nos Pos, Tse Daa Kaan. 

#manual data collection
```{r}
#Population Directly Drawing Water from the Colorado River  
pop_Bodaway_Gap<-1705
pop_Coppermine<-690
pop_LeChee<-1762
pop_Navajo_Mt<-359
pop_chapters_Colorado<- pop_Bodaway_Gap+pop_Coppermine+pop_LeChee+pop_Navajo_Mt

pop_chapters_Colorado #4516

#Population Directly Drawing Water from the San Juan River 
pop_Kayenta<-6035
pop_Oljato<-2729
pop_Shonto<-1645
pop_Beclabito<-724
pop_Upper_Fruitland<-2594
pop_Gadii_Ahi<-430
pop_Mexican_Water<-667
pop_Red_Mesa<-1177
pop_Teec_Nos_Pos<-1454
pop_Tse_Daa_Kaan <-1273

pop_chapters_SJ<-pop_Kayenta+pop_Oljato+pop_Shonto+pop_Beclabito+pop_Upper_Fruitland+pop_Gadii_Ahi+pop_Mexican_Water+pop_Red_Mesa+pop_Teec_Nos_Pos+pop_Tse_Daa_Kaan

pop_chapters_SJ #18728
```

I tried scraping data from the website but it generated empty values. Hence, copy pasting the data was done instead. The code used can be found in the appendix. 

Indian Population drawing water directly from these rivers 
- Colorado River: Estimated to be 4516
- San Juan River: Estimated to be 18728

Navajo people are estimated to use only 7 gallons of water per day. [CITE Source] 
```{r}
#Colorado 
annual_draw_colorado<-4516*7*365
annual_draw_colorado
#San Juan 
annual_draw_SJ<-18728*7*365
annual_draw_SJ

```
cubic meters per second
1 cubic meter = 264.2 gallons 
```{r}
seconds_per_year<-365*24*60*60
water_available_SJ_2024<-500*264.2*seconds_per_year
water_available_SJ_2024
```

# To add to discussion 
7 gallons per day is not enough to meet basic needs. According to the UN WHO, at least 50 liters per day per person is needed, which is about 13.2 gallons. 
Source: https://www.un.org/waterforlifedecade/pdf/human_right_to_water_and_sanitation_media_brief.pdf

# Explaining observed water availability 
1. Drought: https://www.nature.com/articles/s41558-022-01290-z
Rapid intensification of the emerging southwestern North American megadrought in 2020–2021
2. High snowpack from the mountains: The Basin experienced excellent snowpack last winter and above average runoff in 2023.
https://www.azwater.gov/news/articles/2023-12-21-0 


\newpage


## Water Quality
### Question 2: 



\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 

# Appendix
## Data scraping attempt
```{r}
#set scraping website 
Water_URL<- read_html('https://navajoprofile.wind.enavajo.org/')
Water_URL

#scrape the data 
navajo_nation_chapter<-Water_URL%>%
  html_nodes("td:nth-child(3)")%>%
  html_text()
navajo_nation_chapter

indian_population<-Water_URL%>%
  html_nodes("td:nth-child(4)")%>%
  html_text()
indian_population


table<-Water_URL%>%
  html_nodes("tabContent001")%>%
  html_text()
table

#Note: I tried scraping data from the website but it didn't work and it generated empty values.  Here's the code I used. Hence, I decided to copy paste instead. 

```
