---
title: 'Final Project: Navajo Nation Water Quantity and Quality Analysis from 2019-2024'
author: "Samantha White-Murillo, Wynona Curaming, Siyu Dong" 
date: "Spring 2024"
output: 
  pdf_document:
   fig_caption: true
---
\newpage
\tableofcontents
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Working directory
getwd()

# Load your packages
library(dataRetrieval)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)
library(here)
library(ggthemes)
library(pdftables)
library(viridis)
library(rvest)
library(tidycensus)
library(ggplot2)

# Set your ggplot theme
my_theme <- theme_light() +
  theme(plot.title = element_text(face = "bold", size = 12, color = "black"),
        axis.title = element_text(color = "darkblue", )) +
  theme(legend.position = "bottom")

# Load your data sets
## Water Quality
Boron_Site1 <- read.csv("./Raw_data/Boron_S1.csv", stringsAsFactors = TRUE)
Boron_Site2 <- read.csv("./Raw_data/Boron_S2.csv", stringsAsFactors = TRUE)
Magnesium_Site1 <- read.csv("./Raw_data/magnesium_S1.csv", stringsAsFactors = TRUE)
Magnesium_Site2 <- read.csv("./Raw_data/magnesium_S2.csv", stringsAsFactors = TRUE)
Uranium_Site1 <- read.csv("./Raw_data/Uranium_S1.csv", stringsAsFactors = TRUE)
Uranium_Site2 <- read.csv("./Raw_data/Uranium_S2.csv", stringsAsFactors = TRUE)


##Function to summarize dataset structure
summarize_dataset <- function(df, dataset_name) {
  data.frame(
    Dataset = dataset_name,
    Variable = names(df),
    Class = sapply(df, class)
  )
}

##Summarize 6 water quality datasets
summary_Boron_Site1 <- summarize_dataset(Boron_Site1, "Boron_Site1")
summary_Boron_Site2 <- summarize_dataset(Boron_Site2, "Boron_Site2")
summary_Magnesium_Site1 <- summarize_dataset(Magnesium_Site1, "Magnesium_Site1")
summary_Magnesium_Site2 <- summarize_dataset(Magnesium_Site2, "Magnesium_Site2")
summary_Uranium_Site1 <- summarize_dataset(Uranium_Site1, "Uranium_Site1")
summary_Uranium_Site2 <- summarize_dataset(Uranium_Site2, "Uranium_Site2")

##Combine summaries into one table
summary_water_quality <- rbind(summary_Boron_Site1, summary_Boron_Site2,
                               summary_Magnesium_Site1, summary_Magnesium_Site2,
                               summary_Uranium_Site1, summary_Uranium_Site2)

##Print water quality summary table
print(summary_water_quality)


## Water Quantity 
siteNumber <- "09379500"
ChoptankInfo <- readNWISsite(siteNumber)
parameterCd <- "00060"

# Raw daily data:
sanjuan_quantity_df <- readNWISdv(
  siteNumber, parameterCd,
  "2019-04-24", "2024-04-22")

pCode <- readNWISpCode(parameterCd)

siteNumber2<- "09380000"
ChoptankInfo2 <- readNWISsite(siteNumber2)
parameterCd2 <- "00060"

# Raw daily data:
colorado_quantity_df <- readNWISdv(
  siteNumber2, parameterCd2,
  "2019-04-24", "2024-04-22")

pCode2 <- readNWISpCode(parameterCd2)

# Summarize 2 water quantity datasets with the previously built function
summary_sanjuan <- summarize_dataset(sanjuan_quantity_df, "sanjuan_quantity_df")
summary_colorado <- summarize_dataset(colorado_quantity_df, "colorado_quantity_df")

# Combine summaries into one table
summary_water_quantity <- rbind(summary_sanjuan, summary_colorado)

# Print summary table
print(summary_water_quantity)

```

# Rationale and Research Questions
The Navajo Nation has severe water infrastructure deficiencies that impact the health, economy, and welfare of the Navajo people. The lack of adequate domestic and municipal water is the greatest water resource problem facing the Navajo Nation. Approximately 30-40% of the Navajo Nation population does not have access to clean, reliable drinking water  (Navajo Nation Department of Water Resources, 2024). Also, 173 thousand people are affected because drinking water sources are limited, and abandoned uranium mines have caused groundwater contamination in the Nation (Ibid.). In addition, many improvements are needed for other areas of water use, including water for irrigation, livestock, commercial, businesses, health care, schools, and other facilities (SOURCE, n.d.).

In terms of water quantity, climate change in the Southwest will continue to impact water resources problems. The USGS Disaster Risk Assessment Study concluded that a long-term drying trend and decreasing snowpack, superimposed on the regional drought cycles, will magnify water-related impacts in the Navajo Nation and leave the Navajo people increasingly vulnerable (Navajo Nation Department of Water Resources, 2024). 

The Navajo Nation's water sources are mainly the interstate rivers and groundwater, unfortunately they face legal difficulties in getting its interests met within inter-state negotiations on water rights. The Colorado River Compact set in 1922 gives seven states the right to draw water, and are grouped into two, the Upper Basin States (namely Colorado, New Mexico, Utah and Wyoming) and Lower Basin States (namely Arizona, Nevada and California) (Anastasia Hufham, 2024), but the Navajo Nation is not formally included in these agreements. As the post-2026 operations of the river are being negotiated, the Navajo Nation is engaging in persistent legal efforts to be part of the negotiations and have the right to meet its obligation to set aside water for its people.

The current project focuses on analyzing the water quality of the main Colorado River before and after it flows through the reservation over the past 5 years. A five-year analysis period was selected to provide a more current evaluation. Taking into account the tribe’s persistent efforts to compel the federal government to meet its obligations, aiding in the quantification of the tribe’s water rights on the Colorado River and ensuring access to high-quality water, we have chosen to focus our quality analysis on the Colorado River. This decision stems from the river’s current water rights litigation and data availability. This analysis aims to test the following hypotheses: 

> Hypothesis A: The water quality in the Colorado River before and after passing through the Navajo Nation is significantly different.

> Hypothesis B: The water quality in the Colorado River before and after passing through the Navajo Nation has changed significantly in the last 5 years.

On the other hand, measurements have been conducted for the water quantity of the rivers over the last 5 years (2019-2024) to assess water availability in both the Colorado River and the San Juan River. In order to broaden the scope of our analysis of water availability, the San Juan River was also selected, establishing the following hypothesis:

> Hypothesis C: Water availability changed in the last years for the Navajo Nation.

\newpage

# Dataset Information and Methods
The dataset was taken from the USGS Monitoring the rivers of the Nation (https://waterdata.usgs.gov/nwis).

For water quality analysis, data was extracted from the years 2018 to 2023, capturing Uranium, Magnesium, and Boron levels detected at both stations. Those components pose a high risk for human health and the environment at high levels. Uranium is radioactive , and high and prolonged exposure increases risks of kidney diseases (Vengosh, 2024). High concentrations of boron in humans can cause nausea, vomiting, redness of the skin, difficulty swallowing due to ulcers in the throat, and non-bloody diarrhea (U.S. EPA, et al., 2008). Moreover, these components were selected since the data was available for both stations, enabling a comparative analysis. 

Below is the information on the stations worked:
- Station 1 = Colorado River at Lees Ferry, AZ - 09380000
- Station 2 = Colorado River Near Grand Canyon, AZ - 09402500

For water quantity, the data was extracted from years 2019 to 2024 from station 1 as well, extracting data of water flow in cf/s. For the San Juan river, the data was extracted from the gage on site San Juan River near Bluff, UT - FF09379500. 

Wrangling: 
For water quantity, the data was wrangled to have more understandable column names for the discharge. The data was parsed to include only relevant variables, namely, discharge and date. No additional wrangling was needed since the dates and discharge were already in the correct format (Year-mm-dd) and class (Date and number respectively), and the dataset contained no missing data. 

\newpage

# Analysis 

## Water Quantity
#### Question 1: How have water discharge levels changed in the last 5 years at San Juan and Colorado rivers?


```{r Data Wrangling, echo=FALSE}
#Renaming columns to be understandable 
sanjuan_quantity_df<-rename(sanjuan_quantity_df, discharge_cf = X_00060_00003)
colorado_quantity_df<-rename(colorado_quantity_df, discharge_cf = X_00060_00003)

#Parsing to include only relevant columns (discharge and date)
sanjuan_quantity_df<- subset(sanjuan_quantity_df, select=c("Date", "discharge_cf"))
colorado_quantity_df<- subset(colorado_quantity_df, select=c("Date", "discharge_cf"))
```

### Data Visualization

The downward slope measured at both gages shows that the discharge volume for both San Juan and Colorado Rivers have declined in the last five years (Figures 1 and 2). This corroborates with previous years where drought conditions have affected water quantity, posing a continued challenge to water availability for the people in the Navajo Nation and impacting aquatic ecosystems. While both rivers experience seasonal variability, the San Juan River has more frequent periods of high discharge (eg. mid-2023 and mid-2019) compared to the Colorado River. 

```{r Data Visualization San Juan, fig.align='center', fig.cap="Water Levels in San Juan River", message=FALSE, echo=FALSE}
ggplot(data=sanjuan_quantity_df, 
       aes(x=Date, y=discharge_cf)) +
  geom_point() +
  geom_smooth(method="lm") +
  labs(y= "Discharge (cubic feet/second)") +
  my_theme

```

```{r Data Visualization Colorado, echo=FALSE, fig.align='center', fig.cap="Water Levels in Colorado River", message=FALSE}
ggplot(data = colorado_quantity_df, 
       aes(x=Date, y=discharge_cf)) +
  geom_point()+geom_smooth(method="lm") +
  labs(y= "Discharge (cubic feet/second)") +
  my_theme

```

\newpage

### Time Series Analysis

For Colorado River, water levels have declined from 2019 to mid-2022 and then increased to peak at early 2023 (Figure 4). For San Juan River, water levels have declined from 2019 to 2020 and remained steady (at a low level) up till mid-2022. Suddenly, water levels increased to peak at early 2023, and declined since then (Figure 5).

```{r Time Series Analysis San Juan River, echo=FALSE, fig.align='center', fig.cap="Time Series Analysis of San Juan River, 2019-2024"}

# San Juan River 
# Generate time series

f_month <- month(first(sanjuan_quantity_df$Date))
f_year <- year(first(sanjuan_quantity_df$Date))
f_day<- day(first(sanjuan_quantity_df$Date))

l_month <- month(last(sanjuan_quantity_df$Date))
l_year <- year(last(sanjuan_quantity_df$Date))
l_day<- day(last(sanjuan_quantity_df$Date))

sanjuan_quantity_daily_ts <- ts(sanjuan_quantity_df$
                              discharge_cf,
                            start=c(f_year,f_month),
                            frequency=365)

# Decompose

sanjuan_quantity_daily_decomposed <- stl(sanjuan_quantity_daily_ts, s.window = "periodic")
plot(sanjuan_quantity_daily_decomposed)

```

```{r Time Series Analysis Colorado River, echo=FALSE, fig.align='center', fig.cap="Time Series Analysis of Colorado River, 2019-2024"}

# Colorado River  
# Generate time series
f_month1 <- month(first(colorado_quantity_df$Date))
f_year1 <- year(first(colorado_quantity_df$Date))
f_day1<- day(first(colorado_quantity_df$Date))

l_month1 <- month(last(colorado_quantity_df$Date))
l_year1 <- year(last(colorado_quantity_df$Date))
l_day1<- day(last(colorado_quantity_df$Date))

colorado_quantity_daily_ts<- ts(colorado_quantity_df$
                              discharge_cf,
                            start=c(f_year,f_month),
                            frequency=365)

# Decompose
colorado_quantity_daily_decomposed<- stl(colorado_quantity_daily_ts, s.window = "periodic")
plot(colorado_quantity_daily_decomposed)
```

\newpage
### Population Drawing Directly From the Rivers 

We examined each of the chapters under the Western and Northern Agency on whether the San Juan and Colorado River passes through their territory (Navajo Nation Division of Community Development, n.d.). This would allows us to determine how much of the Navajo population drew water directly from each river. 

For Colorado River, the following chapters met this criteria: Bodaway-Gap, Coppermine, LeChee, and Navajo Mountain. For San Juan, the following chapters met this criteria: Kayenta, Oljato, Shonto, Beclabito, Upper Fruitland, Gadii Ahi/to'koi, Mexican Water, Red Mesa, Teec Nos Pos, Tse Daa Kaan. 

The total Indian population in the Navajo Nation estimated to live in the chapters bordering the Colorado River is about 4516 (Navajo Nation Division of Community Development, n.d.). San Juan River has about 18728 people living close-by (Ibid.). 

```{r Manual data collection, echo=FALSE, message=FALSE, include=FALSE}

#Population Directly Drawing Water from the Colorado River  
pop_Bodaway_Gap<-1705
pop_Coppermine<-690
pop_LeChee<-1762
pop_Navajo_Mt<-359
pop_chapters_Colorado<- pop_Bodaway_Gap+pop_Coppermine+pop_LeChee+pop_Navajo_Mt

#Population Directly Drawing Water from the San Juan River 
pop_Kayenta<-6035
pop_Oljato<-2729
pop_Shonto<-1645
pop_Beclabito<-724
pop_Upper_Fruitland<-2594
pop_Gadii_Ahi<-430
pop_Mexican_Water<-667
pop_Red_Mesa<-1177
pop_Teec_Nos_Pos<-1454
pop_Tse_Daa_Kaan <-1273

pop_chapters_SJ<-pop_Kayenta+pop_Oljato+pop_Shonto+pop_Beclabito+pop_Upper_Fruitland+pop_Gadii_Ahi+pop_Mexican_Water+pop_Red_Mesa+pop_Teec_Nos_Pos+pop_Tse_Daa_Kaan

```

*Note: We intended to use scrape data from the website but it generated empty values. Hence, copy pasting the data was done instead. See appendix 1.*

Navajo people are estimated to use only 7 gallons of water per day (Supreme Court of the United States, 2022). Based on this estimate, and the estimate of population drawing water directly from the rivers, we can estimate the gallons of annual water withdrawal by the Navajo Nation. About 11538380 gallons in total are directly withdrawn from the Colorado River by the Navajo Nation, while about 47850040 gallons in total are directly withdrawn by the Navajo Nation from the San Juan River. 

This is a minuscule quantity compared to the water drawn by states in the Lower Basin (Arizona, Nevada and California) for agricultural irrigation, the activity responsible for the largest water use from the Colorado River (Richter et al., 2024). The average water used for agriculture by the Lower Basin from 2000 to 2019 is 6610.06 million cubic meters, which is 1746193109029.7864 gallons (Ibid.). 

As a caveat, we are counting only populations right beside the river, even though with water transport infrastructure, people in chapters far from the river will also be withdrawing, so this estimate of water withdrawing is an underestimate. This estimate applies only for domestic use but since they use water for other activities such as agriculture, the water availability of both rivers is a limiting factor.  

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
seconds_per_year<-365*24*60*60
#water available in periods of minimum water discharge from the last 5 years
water_available_SJ_2024<-500*7.48052*seconds_per_year
water_available_SJ_2024 
water_available_Colorado_2024<-9500*7.48052*seconds_per_year
water_available_Colorado_2024  

```

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#Colorado 
annual_draw_colorado<-4516*7*365
annual_draw_colorado
#San Juan 
annual_draw_SJ<-18728*7*365
annual_draw_SJ

```


```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#conversion: cubic feet per second. 1 cubic feet = 7.48052 gallons 

seconds_per_year<-365*24*60*60
#water available in periods of minimum water discharge from the last 5 years
water_available_SJ_2024<-500*7.48052*seconds_per_year
water_available_SJ_2024 
water_available_Colorado_2024<-9500*7.48052*seconds_per_year
water_available_Colorado_2024  

```

\newpage


## Water Quality
#### Question 2: How has been the water quality of Colorado River before, and after it pass through the reservation in the past 5 years?

```{r Dara Preparation, echo=FALSE, message=FALSE, include=FALSE}

#Change to date format for the Activity_StartDate columns for further analysis
class(Boron_Site1$Activity_StartDate)
class(Boron_Site2$Activity_StartDate)
class(Magnesium_Site1$Activity_StartDate)
class(Magnesium_Site2$Activity_StartDate)
class(Uranium_Site1$Activity_StartDate)
class(Uranium_Site2$Activity_StartDate)

Boron_Site1$Activity_StartDate <- as.Date(Boron_Site1$Activity_StartDate, format = "%Y-%M-%d")
Boron_Site2$Activity_StartDate <- as.Date(Boron_Site2$Activity_StartDate, format = "%Y-%M-%d")
Magnesium_Site1$Activity_StartDate <- as.Date(Magnesium_Site1$Activity_StartDate, format = "%Y-%M-%d")
Magnesium_Site2$Activity_StartDate <- as.Date(Magnesium_Site2$Activity_StartDate, format = "%Y-%M-%d")
Uranium_Site1$Activity_StartDate <- as.Date(Uranium_Site1$Activity_StartDate, format = "%Y-%M-%d")
Uranium_Site2$Activity_StartDate <- as.Date(Uranium_Site2$Activity_StartDate, format = "%Y-%M-%d")

class(Boron_Site1$Activity_StartDate)
class(Boron_Site2$Activity_StartDate)
class(Magnesium_Site1$Activity_StartDate)
class(Magnesium_Site2$Activity_StartDate)
class(Uranium_Site1$Activity_StartDate)
class(Uranium_Site2$Activity_StartDate)

#Merged Datasets respectively for each target contamination
Boron_merged <- bind_rows(Boron_Site1, Boron_Site2)
Magnesium_merged <- bind_rows(Magnesium_Site1, Magnesium_Site2)
Uranium_merged <- bind_rows(Uranium_Site1, Uranium_Site2)

#Unify the research time range. We are going to research for the specific time range from 2018 to 2023. For the contamination of Uranium, its raw dataset have longer time span so here we filter it.
Uranium_filtered <- Uranium_merged %>%
  filter(Activity_StartDate >= as.Date("2018-01-01") & 
         Activity_StartDate <= as.Date("2023-12-31"))

#Generate the annual average concentration of three contamination for both sites
Boron_merged$Year <- lubridate::year(Boron_merged$Activity_StartDate)
Boron_annual_average <- Boron_merged %>%
  group_by(Year, Location_Name) %>%
  summarise(Avg_Concentration = mean(Result_MeasureValue, na.rm = TRUE))

Magnesium_merged$Year <- lubridate::year(Magnesium_merged$Activity_StartDate)
Magnesium_annual_average <- Magnesium_merged %>%
  group_by(Year, Location_Name) %>%
  summarise(Avg_Concentration = mean(Result_MeasureValue, na.rm = TRUE))

Uranium_filtered$Year <- lubridate::year(Uranium_filtered$Activity_StartDate)
Uranium_annual_average <- Uranium_filtered %>%
  group_by(Year, Location_Name) %>%
  summarise(Avg_Concentration = mean(Result_MeasureValue, na.rm = TRUE))


```

### Boron

Mean annual boron concentrations increased on the Colorado River (Figure 5) from 2019 to 2024; nevertheless, the concentrations have not yet exceeded a concerning level for human health according to the recommended maximum levels for drinking water by Canada and The World Health Organization (World Health Organization, 2009; Health Canada, 2020). The US has not set any regulations limiting boron concentrations in water. 

Throughout this 5-year period, boron concentration is lower when entering the tribal territory than when it exits. It reached a maximum concentration of >120 ug/L, which is still considered low, with a general mean concentration along the river of 74.81 ug/L. The statistically significant difference (p-value = 0.002834 < 0.05, df = 8.9481, t = -4.0699) found through a t-test indicates that the Navajo Nation engages in activities that result in boron discharge in the Colorado River. 

```{r Data visualization Boron t-test, echo=FALSE, message=FALSE, include=FALSE}
#t-test
Boron_t_test <- t.test(Boron_Site1$Result_MeasureValue, Boron_Site2$Result_MeasureValue)
print(Boron_t_test)

```

```{r Data visualization Boron, echo=FALSE, fig.align='center', fig.cap="Boron Concentration before and after passing through the Navajo Nation"}

#Compare Boron
ggplot(Boron_annual_average, 
       aes(x = Year, y = Avg_Concentration, color = Location_Name)) +
  geom_point() +
  labs(x = "Observation Year",
       y = "Mean Concentration (ug/L)") +
  scale_color_discrete(name = "Location") +
  my_theme

```

### Magnesium

Mean annual magnesium concentrations increased on the Colorado River (Figure 6) from 2020 to 2023, with a low concentration in 2018 and a significant increase in 2019, reaching up to 22.9 mg/L. Throughout this 5-year period, magnesium concentration is lower upon entering the tribal territory than when it exits, with a significant difference of 93.61% between both sides, resulting in a mean concentration along the river of 21.55 mg/L. This element is one of the most common elements in natural resources and is essential for human nutrition. The levels found in the water source are typical in water bodies and do not represent a risk to human and natural health (U.S. EPA, 1984).

```{r Data visualization Magnesium t-test, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#t-test
Magnesium_t_test <- t.test(Magnesium_Site1$Result_MeasureValue, Magnesium_Site2$Result_MeasureValue)
print(Magnesium_t_test)

```

```{r Data visualization Magnesium, echo=FALSE, fig.align='center', fig.cap="Magnesium Concentration before and after passing through the Navajo Nation"}
#Compare Magnesium
ggplot(Magnesium_annual_average, 
       aes(x = Year, y = Avg_Concentration, color = Location_Name)) +
  geom_point() +
  labs(x = "Observation Year",
       y = "Mean Concentration (ug/L)") +
  scale_color_discrete(name = "Location")+
  my_theme

```

### Uranium

Mean annual uranium concentrations have increased on the Colorado River (see Figure 5) from 2019 to 2024. Nevertheless, the concentrations have not yet exceeded concerning levels for human health according to the recommended maximum levels for drinking water by The World Health Organization and the EPA. However, the goal concentrations are zero (World Health Organization, 2009; US EPA, 2024). Considering that the water used from the Colorado River is for direct human consumption and other activities, accumulation of this element can potentially generate radiological toxicity. The low concentrations of this component at this location indicate a positive environmental stage considering the historical mining in the Navajo territories (Doug Brugge, Rob Goble, 2002).

Throughout this 5-year period, uranium concentration is lower when entering the tribe territory than when it exits (except in 2021 and 2023), reaching a maximum concentration of 5.80 ug/L, with a general mean concentration along the river of 3.47 ug/L. The statistically significant difference (p-value = 8.993e-5) indicates a difference in uranium measurements between the entrance to the reservation and the exit of the water concentrations of the Colorado River, suggesting that there are activities or naturally occurring uranium discharges.

```{r Data visualization Uranium t-test, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#t-test
Uranium_t_test <- t.test(Uranium_Site1$Result_MeasureValue, Uranium_Site2$Result_MeasureValue)
print(Uranium_t_test)

```


```{r Data visualization Uranium, echo=FALSE, fig.align='center', fig.cap="Uranium Concentration before and after passing through the Navajo Nation"}

#Compare Uranium
ggplot(Uranium_annual_average, 
       aes(x = Year, y = Avg_Concentration, color = Location_Name)) +
  geom_point() +
  labs(x = "Observation Year",
       y = "Mean Concentration (ug/L)") +
  scale_color_discrete(name = "Location")+
  my_theme

```

\newpage

# Summary and Conclusions
boron, magnecium ratio

\newpage

# References

Anastasia Hufham. (2024). Utah says it shouldn’t have to take Colorado River cuts. Other states say it needs to. The Salt Lake Tribune. https://www.sltrib.com/news/environment/2024/03/06/utah-says-it-shouldnt-have-take/

Doug Brugge, Rob Goble. (2002). The History of Uranium Mining and the Navajo People. Public health then and now. https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3222290/pdf/0921410.pdf

Health Canada. (2020). Boron in drinking water. In Guidelines for Canadian Drinking Water Quality: Guideline Technical Document [Report]. https://www.canada.ca/content/dam/hc-sc/documents/programs/consultation-boron-drinking-water/document/boron-gtd-consultation-09-12-2019-eng.pdf

National Primary Drinking Water Regulations | US EPA. (2024, January 2). US EPA. https://www.epa.gov/ground-water-and-drinking-water/national-primary-drinking-water-regulations

Navajo Nation Division of Community Development (n.d.). Navajo Nation Profile. Navajo Nation WIND. Retrieved April 28, 2024, from https://navajoprofile.wind.enavajo.org/

Navajo Nation Water Rights Comission. (2024, 3 22). Navajo Nation Water Rights Comission 
webpage. Retrieved from https://nnwrc.navajo-nsn.gov/

Richter, B. D., Lamsal, G., Marston, L., Dhakal, S., Sangha, L. S., Rushforth, R. R., Wei, D., Ruddell, B. L., Davis, K. F., Hernandez-Cruz, A., Sandoval-Solis, S., & Schmidt, J. C. (2024). New water accounting reveals why the Colorado River no longer reaches the sea. Communications Earth & Environment, 5(1), 134. https://doi.org/10.1038/s43247-024-01291-0

SOURCE. (n.d.). Los hidropaneles SOURCE llevan agua a la Nación Navajo. Retrieved from https://www.source.co/resources/case-studies/los-hidropaneles-source-llevan-agua-a-lanacion-navajo/?lang=es

Supreme Court of the United States (2022, October 1). Arizona et al. V. Navajo Nation et al. Retrieved April 28, 2024, from https://www.supremecourt.gov/opinions/22pdf/21-1484_aplc.pdf

The Colorado River: Heading into 2024 with hope for a more stable system | Arizona Department of Water Resources. (2023, December 21). https://www.azwater.gov/news/articles/2023-12-21-0

U.S. EPA, Health and Ecological Criteria Division (HECD), O. of S. and T. (OST), OW, Office of Water (OW), & Office of Groundwater/Drinking Water (OGWDW), O. (2008). Drinking Water Health Advisory for Boron and Compounds. https://www.epa.gov/sites/default/files/2014-09/documents/summary_document_from_the_ha_for_boron.pdf

U.S. EPA. (1984). Ambient surface water quality standards documentation, Magnesium. https://www.epa.gov/sites/default/files/2015-06/documents/ny_hh_396_w_12201984.pdf

UN-Water Decade Programme on Advocacy and Communication and Water Supply and Sanitation Collaborative Council. (2010). The human right to water and sanitation today. In Media Brief. https://www.un.org/waterforlifedecade/pdf/human_right_to_water_and_sanitation_media_brief.pdf

Williams, A. P., Cook, B. I., & Smerdon, J. E. (2022). Rapid intensification of the emerging southwestern North American megadrought in 2020–2021. Nature Climate Change, 12(3), 232–234. https://doi.org/10.1038/s41558-022-01290-z

World Health Organization. (2009). Boron. Chemical Fact Sheets. https://www.who.int/docs/default-source/wash-documents/wash-chemicals/boron-chemical-fact-sheet.pdf?sfvrsn=a936384f_4

\newpage

# Appendix
## Appendix 1
```{r Data scraping attempt}
#set scraping website 
Water_URL<- read_html('https://navajoprofile.wind.enavajo.org/')
Water_URL

#scrape the data 
navajo_nation_chapter<-Water_URL%>%
  html_nodes("td:nth-child(3)")%>%
  html_text()
navajo_nation_chapter

indian_population<-Water_URL%>%
  html_nodes("td:nth-child(4)")%>%
  html_text()
indian_population


table<-Water_URL%>%
  html_nodes("tabContent001")%>%
  html_text()
table

```
