---
title: "Water Qualtiy Analysis"
author: "Wynona Curaming"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
## Set Up
```{r, setup}
getwd()
library(dataRetrieval)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)
library(here)
library(ggthemes)
```

## Importing Datasets
```{r}
siteNumber <- "09379500"
ChoptankInfo <- readNWISsite(siteNumber)
parameterCd <- "00060"

# Raw daily data:
sanjuan_quantity_df <- readNWISdv(
  siteNumber, parameterCd,
  "2019-04-24", "2024-04-22"
)

pCode <- readNWISpCode(parameterCd)

```

```{r}
summary(sanjuan_quantity_df)
```

```{r}
siteNumber2<- "09380000"
ChoptankInfo2 <- readNWISsite(siteNumber2)
parameterCd2 <- "00060"

# Raw daily data:
colorado_quantity_df <- readNWISdv(
  siteNumber2, parameterCd2,
  "2019-04-24", "2024-04-22"
)

pCode2 <- readNWISpCode(parameterCd2)
```

```{r}
summary(colorado_quantity_df)
```
## Data Wrangling
```{r}
#Renaming columns to be understandable 
sanjuan_quantity_df<-rename(sanjuan_quantity_df, discharge_cf = X_00060_00003)
colorado_quantity_df<-rename(colorado_quantity_df, discharge_cf = X_00060_00003)

```

## Data Visualization 
```{r, data exploration}
ggplot(data=sanjuan_quantity_df, aes(x=Date, y=discharge_cf))+geom_point()+geom_smooth(method="lm")
ggplot(data=colorado_quantity_df, aes(x=Date, y=discharge_cf))+geom_point()+geom_smooth(method="lm")

discharge_sanjuan<-sanjuan_quantity_df$discharge_cf
plot(discharge_sanjuan)
```
> The downward slope measured at both gages suggests that discharge volume for both San Juan and Colorado Rivers have declined in the last five years. San Juan appears to have more frequent periods of high discharge than usual (eg. mid-2023 and mid-2019) compared to Colorado. 

## Time Series Analysis 
Study question: Have water discharge levels changed in the last 5 years at San Juan and Colorado rivers?

We don't have any missing data so we don't need to do linear interpolation. 

```{r}
# San Juan River 
# Generate time series
f_month <- month(first(sanjuan_quantity_df$Date))
f_year <- year(first(sanjuan_quantity_df$Date))
f_day<- day(first(sanjuan_quantity_df$Date))

l_month <- month(last(sanjuan_quantity_df$Date))
l_year <- year(last(sanjuan_quantity_df$Date))
l_day<- day(last(sanjuan_quantity_df$Date))

sanjuan_quantity_daily_ts<- ts(sanjuan_quantity_df$
                              discharge_cf,
                            start=c(f_year,f_month),
                            frequency=365)

# Decompose

sanjuan_quantity_daily_decomposed<- stl(sanjuan_quantity_daily_ts, s.window = "periodic")
plot(sanjuan_quantity_daily_decomposed)

```

```{r}
# Colorado River  
# Generate time series
f_month1 <- month(first(colorado_quantity_df$Date))
f_year1 <- year(first(colorado_quantity_df$Date))
f_day1<- day(first(colorado_quantity_df$Date))

l_month1 <- month(last(colorado_quantity_df$Date))
l_year1 <- year(last(colorado_quantity_df$Date))
l_day1<- day(last(colorado_quantity_df$Date))

colorado_quantity_daily_ts<- ts(colorado_quantity_df$
                              discharge_cf,
                            start=c(f_year,f_month),
                            frequency=365)

# Decompose
colorado_quantity_daily_decomposed<- stl(colorado_quantity_daily_ts, s.window = "periodic")
plot(colorado_quantity_daily_decomposed)

```



check how much water the Navajo people withdraw or need?
check population living near the river-> est how much they would need 
potential explanation- agriculture
